Gusto: a toolkit for compatible finite element dynamical cores

### 2024-07-25 source/paper.tex
add captions to listings and make a few changes

\section{Form Manipulation Language}
\label{sec: FML}
The discussion of the time discretisation so far has been general i.e.
not specific to any particular equation set. This is also true of the
implementation of the timestepping schemes. This enables flexibility
and reproducibility (**elaborate). This functionality is enabled
through the form manipulation language (FML) which is used to
associate a label to each term in the residual form defined in the
equation class. The timestepping methods are then able to reason about
which manipulations to apply to which forms by applying filters that
check for the presence of a label and, optionally, its value. Here we
will describe the objects required to define labels and filters before
illustrating this with a simple example(s?).

The finite element discretisation of each term in an equation is
represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
finite element discretisation of the term and a dictionary of
key-value pairs corresponding to the FML \texttt{Label} objects that
apply to this term. An FML \texttt{Label} object has a name and an
optional value, defaulting to \texttt{True} to indicate the presence
of the label. Labels can also be used to tag parts of the form. For
example, the \texttt{subject} label is used to tag the
\texttt{Function} in the \texttt{ufl.Form} that represents the
prognostic variable. These operations are composable, so more than one
label can be attached to a term. The end result of labelling the terms
in the residual is a \texttt{LabelledForm} object that holds a list of
terms, pairing individual \texttt{ufl.Form} objects with FML
labels. Labelled forms can be added together, multiplied by floats and
have additional terms added. More complex manipulations are enabled by
the \texttt{label\_map} method that provides a way of filtering the
terms in the form based on their labels, applying different mappings
to terms that pass or fail the conditions defined in the filter. A
common operation might be to filter for terms that have a certain
label, such as \texttt{transport}, drop terms that do not have this
label and multiply those that do by the timestep. We shall demonstrate
this in the following example.

\JScomment{describe methods has label and get?}

@@

\lstinputlisting[caption=blah, label={lst:define-residual}, language=python, firstline=24, lastline=31]{code/advection_diffusion_splitting_example.py}

@@

\lstinputlisting[caption=blah, label={lst:define-explicit}, language=python, firstline=38, lastline=53]{code/advection_diffusion_splitting_example.py}

@@

\lstinputlisting[caption=blah, label={lst:define-implicit}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}

The final part of the code is the time loop which calls the solve
method of the explicit and then the impicit solvers defined above, see
listing \ref{lst:timeloop}

>>>>

\section{Form Manipulation Language}
\label{sec: FML}
The discussion of the time discretisation so far has been general i.e.
not specific to any particular equation set. This is also true of the
implementation of the timestepping schemes, resulting in modular code
where the equations are defined separately from the timestepping
scheme enabling flexibility in combining different equations with
different timestepping schemes. This powerful functionality is enabled
through the form manipulation language (FML) which is used to
associate a label to each term in the residual form defined in the
equation class. The timestepping methods are then able to reason about
which manipulations to apply to which forms by applying filters that
check for the presence of a label and, optionally, its value. Here we
will describe the objects required to define labels and filters before
illustrating this with a simple example(s?).

The finite element discretisation of each term in an equation is
represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
Label object which tags a ufl.Form to become a Term, with the
intention that the Term represents a term in the equation.

FML defines a
\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
finite element discretisation of the term and a dictionary of
key-value pairs corresponding to the FML \texttt{Label} objects that
apply to this term. An FML \texttt{Label} object has a name and an
optional value, defaulting to \texttt{True} to indicate the presence
of the label. Labels can also be used to tag parts of the form. For
example, the \texttt{subject} label is used to tag the
\texttt{Function} in the \texttt{ufl.Form} that represents the
prognostic variable. These operations are composable, so more than one
label can be attached to a term. The end result of labelling the terms
in the residual is a \texttt{LabelledForm} object that holds a list of
terms, pairing individual \texttt{ufl.Form} objects with FML
labels. Labelled forms can be added together, multiplied by floats and
have additional terms added. More complex manipulations are enabled by
the \texttt{label\_map} method that provides a way of filtering the
terms in the form based on their labels, applying different mappings
to terms that pass or fail the conditions defined in the filter. A
common operation might be to filter for terms that have a certain
label, such as \texttt{transport}, drop terms that do not have this
label and multiply those that do by the timestep. We shall demonstrate
this in the following example.

@@

\lstinputlisting[caption={This code defines and labels the residual corresponding to equation \ref{eq:weak-advdiff}}, label={lst:define-residual}, language=python, firstline=24, lastline=31]{code/advection_diffusion_splitting_example.py}

@@

\lstinputlisting[caption={This code defines the solver for the explicit step of the timestepping method i.e. equation \ref{eq:explicit-advdiff}. \texttt{qn} and \texttt{qstar} are Firedrake \texttt{Functions} with \texttt{qn} holding the values of $q$ at time step $n$ and \texttt{qstar} holding the values after the explicit solve.}, label={lst:define-explicit}, language=python, firstline=40, lastline=55]{code/advection_diffusion_splitting_example.py}

@@

\lstinputlisting[caption={This code defines the solver for the implicit step of the timestepping method i.e. equation \ref{eq:implicit-advdiff}. \texttt{qstar} and \texttt{qnp1} are Firedrake \texttt{Functions} with \texttt{qstar} holding the values after the explicit solve and \texttt{qnp1} holding the values of $q$ at after the implicit solve.}, label={lst:define-implicit}, language=python, firstline=57, lastline=71]{code/advection_diffusion_splitting_example.py}

The final part of the code is the time loop which calls the solve
method of first the explicit solver and then the implicit solver, see
listing \ref{lst:timeloop}. In practice, the code in listings
\ref{lst:define-explicit}-\ref{lst:define-implicit} would be in a
function that could be applied to any residual labeled with the
appropriate labels.
